---
title: "Estimating stranding probability"
description: "Workflow for estimating coastal stranding probability from particle trajectories and coastline grid cells."
format: gfm
keep-md: true
output: "../Results/3_Estimating_stranding_prob.md"
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

This analysis estimates the probability that a particle trajectory reaches the coastline within a set of coastal grid cells. It combines particle trajectory data with a spatial grid and assigns each trajectory point to the nearest coastline cell when it falls inside a polygon.

## 1. Load required libraries

```{r setup}
library(tidyverse)
library(sf)
library(parallel)
library(ggplot2)
```

## 2. Load trajectory data

```{r load-trajectory-data}
traj1 <- read.csv("../Results/Goa_fwd_op_p1.csv")
traj2 <- read.csv("../Results/Goa_fwd_op_p2.csv")
traj <- bind_rows(traj1, traj2)
rm(traj1, traj2)
```

## 3. Define coastline grid cells

The coastal cells used here are fixed as a study-specific coastline mask. The code below preserves the original values exactly as defined in the original script.

```{r define-coast-cells}
coast_cells <- c(
  290,338,385,386,433,434,482,435, # Maharashtra
  483,484,531,532,533,581,582,630,631, # Goa
  678,632,679,727,728,776,777,778,779,780,827,828, # Goa
  875,876,877,924,534,726,729, # Goa
  925,972,973,974,927,975,1022,1069,1116 # Karnataka
) # coastline cells
```

## 4. Load and join the spatial grid

The full grid is read from the shapefile and then joined with a CSV containing distance metadata for each cell.

```{r load-grid-data}
grid_full <- read_sf("../Data/Shp files/Grid_ecol_refined.shp", crs = 4326)
grid_dat <- read.csv("../Data/Shp files/Grid_ecol_refined.csv")
grid_full <- grid_full |> left_join(grid_dat[, c("id", "distance")], by = "id")

# Restrict to coastal cells for the stranding analysis
grid <- grid_full |> filter(id %in% coast_cells)
```

## 5. Check whether particle points fall inside a polygon

This helper function identifies whether each trajectory point intersects one coastal polygon. It returns the matched cell ID if the point falls inside exactly one polygon.

```{r polygon-check-function}
is_in_polygon <- function(ipl, grid) {
  pnts <- data.frame(
    x = ipl$feature_x,
    y = ipl$feature_y
  )

  pnts_sf <- sf::st_as_sf(pnts, coords = c("x", "y"), crs = st_crs(grid))
  intersections <- st_intersects(pnts_sf, grid)
  is_inside <- lengths(intersections) == 1
  rec_id <- vapply(intersections, function(x) if (length(x) == 1) x else NA_integer_, integer(1))

  data.frame(
    yy = is_inside,
    rec_id = rec_id
  )
}
```

## 6. Filter trajectories to the study area near the coast

The analysis is limited to trajectory points that fall within a rectangular coastal window covering the Goa coastline.

```{r filter-near-coast}
traj <- traj[
  which(
    (traj$feature_x > 73.650) &
      (traj$feature_y < 15.76) &
      (traj$feature_y > 14.89)
  ),
]
```

## 7. Identify coastal intersections in parallel

For efficiency, the trajectory points are processed in parallel, and the output is combined back into a single data frame.

```{r parallel-stranding-check}
traj_list <- lapply(seq_len(nrow(traj)), function(i) {
  traj[i, c("feature_x", "feature_y")]
})

cl <- makeCluster(12)
clusterExport(cl, c("is_in_polygon", "grid"))
op_pnts <- parLapply(cl, traj_list, function(pt) is_in_polygon(pt, grid))
stopCluster(cl)

op_pnts <- do.call(rbind, op_pnts)
```

## 8. Attach stranding flags to the original trajectory table

A particle is counted as stranded when it intersects a coastline cell exactly once.

```{r attach-stranding-results}
traj$strand <- op_pnts$yy
traj$strand_id <- op_pnts$rec_id
```

## 9. Estimate monthly stranding probability by grid cell

The workflow first converts each trajectory point into a binary indicator of stranding, then groups the data by grid cell, month, and year. A value of 1 indicates at least one stranding event in that group; values are capped and normalized to a probability scale between 0 and 1.

```{r compute-grid-probability}
part_strand_prob <- traj |>
  group_by(id, strand_id, time_slot, month, year) |>
  summarise(count = sum(strand), .groups = "drop") |>
  mutate(count = ifelse(count > 0, 1, 0))

part_strand_prob_from_grid <- part_strand_prob |>
  group_by(id, month, year) |>
  summarise(prob = sum(count), .groups = "drop") |>
  mutate(prob = ifelse(prob > 3, 3, prob), prob = prob / 3) |>
  left_join(grid_full[, c("id", "distance")], by = "id") |>
  group_by(id, distance) |>
  summarise(prob = mean(prob), .groups = "drop")
```

## 10. Save and map the probability surface

```{r save-and-visualize}
# write.csv(part_strand_prob_from_grid, "../Results/str_prob_from_grids.csv", row.names = FALSE)

grid_full |>
  left_join(part_strand_prob_from_grid, by = c("id" = "id")) |>
  ggplot() +
  geom_sf(aes(fill = prob)) +
  scale_fill_viridis_c(option = "plasma", na.value = "lightgrey") +
  theme_minimal() +
  labs(fill = "Stranding probability") +
  ggtitle("Stranding probability from each grid cell")

# ggsave("../Results/Graphs/str_prob_from_grids.jpg", width = 8, height = 6)
```

## 11. Estimate monthly stranding probability across all grids

This summary reports the mean monthly stranding probability across all grid cells.

```{r monthly-summary-all-grids}
part_strand_prob_monthly <- part_strand_prob |>
  group_by(id, month, year) |>
  summarise(prob = sum(count), .groups = "drop") |>
  mutate(prob = ifelse(prob > 3, 3, prob), prob = prob / 3) |>
  group_by(month, year) |>
  summarise(
    mean_prob = mean(prob),
    se = sd(prob) / sqrt(length(prob)),
    .groups = "drop"
  )

# write.csv(part_strand_prob_monthly, "../Results/str_prob_month_year.csv", row.names = FALSE)
```

## 12. Restrict analysis to coastal cells within a short distance of shore

This section focuses on grid cells within 10 km of the coast, which is the biologically relevant coastal band for this species.

```{r coastal-band-summary}
part_strand_prob_nearshore <- part_strand_prob |>
  group_by(id, month, year) |>
  summarise(prob = sum(count), .groups = "drop") |>
  mutate(prob = ifelse(prob > 3, 3, prob), prob = prob / 3) |>
  left_join(grid_full[, c("id", "distance")], by = "id") |>
  filter(distance < 10000)

part_strand_prob_monthly_nearshore <- part_strand_prob_nearshore |>
  group_by(month, year) |>
  summarise(
    mean_prob = mean(prob),
    se = sd(prob) / sqrt(length(prob)),
    .groups = "drop"
  )

# write.csv(part_strand_prob_monthly_nearshore, "../Results/str_prob_month_year_g10km.csv", row.names = FALSE)
```

## 13. Average stranding probability across months

The script then summarizes the mean probability by month and creates a bar plot with standard error bars.

```{r monthly-mean-plot}
part_strand_prob_from_grid_months <- part_strand_prob_nearshore |>
  select(-geometry) |>
  group_by(month) |>
  summarise(
    mean_prob = mean(prob),
    se = sd(prob) / sqrt(length(prob)),
    ci5 = quantile(prob, 0.05),
    ci95 = quantile(prob, 0.95),
    .groups = "drop"
  )

fig_strp <- part_strand_prob_from_grid_months |>
  ggplot(aes(x = as.factor(month), y = mean_prob)) +
  geom_bar(stat = "identity", fill = "gray") +
  geom_errorbar(
    aes(ymin = mean_prob - 1.95 * se, ymax = mean_prob + 1.95 * se),
    width = 0.2
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 14)
  ) +
  labs(y = "Stranding probability\n(mean ± 95% CI)") +
  annotate(
    "text",
    x = 1,
    y = Inf,
    label = "a)",
    hjust = 1.1,
    vjust = 1.5,
    size = 6
  )

fig_strp
```

## 14. Optional combined figure

The original script references a separate population figure (`fig_pop`) from a different analysis file. This step is optional and only works if that object is already loaded in the session.

```{r combined-figure, eval=FALSE}
# Requires fig_pop to be available in the environment from Density_Distance.R
# ggpubr::ggarrange(fig_strp, fig_pop, ncol = 1)
# ggsave("../Results/Graphs/str_prob_pop_distance_CI.jpg", width = 6, height = 8)
```

## 15. Additional exploratory output

The last section computes a relative stranding probability for each destination grid cell by month. This is useful as a comparative index rather than an absolute probability estimate.

```{r relative-probability-to-grid, eval=FALSE}
part_strand_prob_to_grid <- part_strand_prob |>
  group_by(strand_id, month, year) |>
  summarise(prob = sum(count), .groups = "drop") |>
  mutate(prob = prob / max(prob))
```

This workflow estimates a spatially explicit coastal stranding risk surface from trajectory data. The key assumptions are that the coastline mask is appropriate for the study region and that the selected coastal cells capture the stranding process of interest.
