---
title: "Mapping mortality"
description: "Spatial mapping of inferred cetacean mortality risk from stranding records"
format: gfm
keep-md: true
output: "../Results/5_Mapping_mortality.md"
execute:
  warning: false
  message: false
---

# Overview

This code maps spatial mortality risk inferred from the stranding dataset. The workflow assigns each record to a spatial grid cell, applies a set of biological and distance-based filters, and then converts the retained records into a mortality index that can be visualised as a spatial surface across Goa.

The analysis proceeds in four parts:

1. read and filter the raw stranding records,
2. assign each record to its corresponding grid cell,
3. apply the mortality filters and calculate the spatial index, and
4. generate the final map and output table.

## 1. Data import and screening

The first step reads the cleaned review dataset and restricts it to the focal taxon (Humpback dolphins) and records with valid location information. This keeps the analysis limited to the subset that can support a spatial mortality estimate.

```{r}
library(tidyverse)
library(sf)

strand_records <- read.csv("../Results/Goa_rev.csv") |>
  arrange(id, desc(Year), desc(Month), desc(Day)) |>
  filter(flag == "FALSE" | flag == "" | is.na(flag)) |>
  filter(Species_co == "Humpback dolphin") |>
  filter(!is.na(lat))

goa_grid <- st_read("../Data/Shp files/Grid_ecol_refined.shp", quiet = TRUE)
goa_coast <- st_read("../Data/Shp files/Goa_coast.shp", quiet = TRUE)
```

### Result

The filtered dataset contains the set of records that are appropriate for spatial mortality mapping, while the shoreline and coastal-grid layers define the spatial domain used for aggregation.

## 2. Grid assignment logic

Each stranding record is assigned to its corresponding grid cell by checking whether its point falls inside a polygon of the Goa grid.

```{r}
is_in_polygon <- function(point_df, grid_sf) {
  point_sf <- st_as_sf(
    point_df,
    coords = c("lon", "lat"),
    crs = st_crs(grid_sf)
  )

  cell_intersections <- st_intersects(point_sf, grid_sf)
  cell_match <- sapply(cell_intersections, function(x) {
    if (length(x) == 1) x[1] else NA_integer_
  })

  data.frame(
    yy = lengths(cell_intersections) == 1,
    rec_id = cell_match,
    stringsAsFactors = FALSE
  )
}

assign_grid_id <- function(dat, grid_sf) {
  point_assignments <- map_dfr(seq_len(nrow(dat)), function(i) {
    pt <- dat[i, c("lon", "lat")]
    is_in_polygon(pt, grid_sf)
  })

  dat$strand_id <- grid_sf$id[point_assignments$rec_id]
  dat
}
```

The function is applied over all spatial coordinates and returns the grid-cell index for records that fall inside the study grid.

## 3. Distance-to-coast and raw spatial view

The code computes the nearest distance from each record to the coastline and then displays the raw point locations over the grid and coast.

```{r}
point_sf <- st_as_sf(
  data.frame(
    x = strand_records$lon,
    y = strand_records$lat
  ),
  coords = c("x", "y"),
  crs = 4326
)

strand_records <- strand_records |>
  mutate(
    dist_to_coast = as.numeric(st_distance(point_sf, goa_coast))
  )

strand_records <- assign_grid_id(strand_records, goa_grid)

ggplot() +
  geom_sf(data = goa_grid, fill = "white") +
  geom_point(
    data = strand_records,
    mapping = aes(x = lon, y = lat, colour = as.factor(Month)),
    size = 2,
    alpha = 0.7
  ) +
  geom_sf(data = goa_coast, alpha = 0.3) +
  theme_minimal() +
  labs(
    x = "Longitude",
    y = "Latitude",
    colour = "Month"
  )
```

### Result

This view provides the spatial context for the records before filtering; it helps check whether the stranding observations are concentrated near particular coastal areas or across the full grid.

## 4. Mortality index calculation

The mortality index is constructed from records that satisfy the timing filter and remain within the spatial range relevant to the study. Each record contributes a weight of $1/n$ for its unique carcass ID, so that multiple observations linked to the same event do not inflate a single mortality signal. Only grids that are <5 km from the coast are retained and advection time is limited to the maximum post-mortem interval (PMI) for the species.

```{r}
filtered_records <- strand_records |>
  group_by(id) |>
  filter(adv_time_days <= PMI_max) |>
  filter(dist_to_coast < 5000)

filtered_records <- filtered_records |>
  group_by(id) |>
  summarise(n = n(), .groups = "drop") |>
  left_join(filtered_records, by = "id") |>
  group_by(id) |>
  mutate(MI = 1 / n)

cell_mortality <- filtered_records |>
  group_by(strand_id) |>
  summarise(mortality = sum(MI, na.rm = TRUE), .groups = "drop")

mortality_map_data <- goa_grid |>
  left_join(cell_mortality, by = c("id" = "strand_id")) |>
  filter(!is.na(mortality))
```

### Result

The aggregated grid table gives a mortality index for each cell, representing the summed contribution from all retained records within that cell.

## 5. Spatial mortality map

The final map visualises the mortality index across the grid and highlights strata where the inferred mortality risk is greatest.

```{r}
plot_mortality_map <- ggplot() +
  geom_sf(data = mortality_map_data, aes(fill = mortality)) +
  scale_fill_viridis_c(option = "plasma", na.value = "lightgrey") +
  geom_sf(data = goa_coast, alpha = 0.3) +
  geom_point(
    data = filtered_records |> filter(adv_time_days == 0),
    mapping = aes(x = lon, y = lat),
    colour = "red",
    size = 1,
    alpha = 0.7
  ) +
  theme_minimal() +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "Mortality index"
  ) +
  ggtitle("Spatial pattern of inferred mortality risk")

plot_mortality_map
```

### Result

The map identifies the coastal cells with the largest summed mortality index. These cells represent the areas where stranding evidence is most concentrated after applying the filters, and they are the most likely spatial hotspots of latent mortality risk.

## 6. Output

```{r}
#| eval: false
write.csv(
  mortality_map_data,
  "../Results/Goa_rev_IOHD_maxPMI_5km.csv",
  row.names = FALSE
)
```

This final file stores the mapped mortality surface in a format that can be used in downstream summaries or figure generation.
